---
title: Lecture 2 - Importing data and more basics
author: Prof. Alexandra Chouldechova
date: January 19, 2017
font-family: Gill Sans
---


###Agenda

- Wrapping up vector indexing

- Importing data

- Simple summaries of categorical and continuous data

- Coding style

- Homework 1 assigned, grading rubric

- Lab 2

###Wrapping up vector indexing

Let's go back to where we left off in the [lecture 1 slides](http://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture01/lecture01-94842-Rpres.html#/46).  


###Importing data

- Start with survey results from "Homework 0"

- To import tabular data into R, we use the **`read.table()`** command

```{r}
survey <- read.table("http://www.andrew.cmu.edu/user/achoulde/94842/data/survey_data.csv", header=TRUE, sep=",") 
```

- Let's parse this command one component at a time 
  - The data is in a file called `survey_data.csv`, which is an online file 
  - The file contains a `header` as its first row
  - The csv format means that the data is comma-separated, so `sep=","`

- Could've also used `read.csv()`, which is just `read.table()` with the preset `sep=","`

###Exploring the data
- R imports data into a **`data.frame`** object
```{r}
class(survey)
```

- To view the first few rows of the data, use **`head()`**
```{r}
head(survey, 3)
```

- `head(data.frame, n)` returns the first `n` rows of the data frame

- In the Console, you can also use `View(survey)` to get a spreadsheet view

###Simple summary
- Use the `str()` function to get a simple summary of your data set

```{r}
str(survey)
```
<br>

- This says that TVhours is a numeric variable, while all the rest are factors (categorical)

###Another simple summary

```{r}
summary(survey)
```

###Data frame basics

- We will talk more about lists and data frames next week, so only give an intro here

- To see what an R object is made up of, you can use `attributes()`

```{r}
attributes(survey)
```

> An R **data frame** is a *list* whose columns you can refer to by *name* or *index*

###Data frame dimensions
- We can use `nrow()` and `ncol` to determine the number of survey responses and the number of survey questions
```{r}
nrow(survey) # Number of rows (responses)
ncol(survey) # Number of columns (questions)
```
- When writing reports, you will often want to say how large your sample size was
- To do this *inline*, use the syntax:
```{r, eval=FALSE}
`r nrow(survey)`
```
- This allows us to write "`r nrow(survey)` students responded to the survey", and have the number displayed
  automatically change when `nrow(survey)` changes.
  
###Inline code chunks example

- Here's a more complex example of inline code use.

```{r, eval=FALSE}
We collected data on `r ncol(survey)` survey questions from `r nrow(survey)` respondents.  Respondents represented `r length(unique(survey[["Program"]]))` CMU programs.  `r sum(survey[["Program"]] == "PPM")` of the repondents were from PPM.
```

- Which results in

> We collected data on `r ncol(survey)` survey questions from `r nrow(survey)` respondents.  Respondents represented `r length(unique(survey[["Program"]]))` CMU programs.  `r sum(survey[["Program"]] == "PPM")` of the repondents were from PPM.

<br>
- **IMPORTANT**: You are expected to use inline code chunks instead of copying and pasting output whenever possible.

###Indexing data frames

- There are many different ways of indexing the same piece of a data frame
  - Each vector below contains `r nrow(survey)` entries.  For display purposes, the settings have been adjusted so that only the first 22 are shown below
```{r, eval = FALSE}
survey[["Program"]]  # "Program" element
```
```{r, echo = FALSE}
head(survey[["Program"]], 22)
```

```{r, eval = FALSE}
survey$Program # "Program" element
```
```{r, echo = FALSE}
head(survey$Program, 22) # "Program" element
```

```{r, eval = FALSE}
survey[,1] # Data from 1st column
```
```{r, echo = FALSE}
head(survey[,1], 22) # "Program" element
```

###More indexing
- Note that single brackets and double brackets have different effects
```{r, eval = FALSE}
survey[["Program"]]  # Returns the Program column as a vector
```
```{r, echo = FALSE}
head(survey[["Program"]], 22)
```
```{r}
survey["Program"] # sub-data-frame containing only "Program"
```


###Bar plot (categorical data)
<center>
```{r, eval = FALSE}
plot(survey[["Program"]])
```
```{r, fig.width=14, fig.height=10, echo = FALSE}
plot(survey[["Program"]],
      cex.lab = 3.5, cex.axis = 3.5, cex.main = 3.5, cex.names = 3.5)
```
</center>

###Histogram (continuous data)
<center>
```{r, eval = FALSE}
hist(survey$TVhours, col="lightgray")
```
```{r, fig.width=16, fig.height=12, echo = FALSE}
# This is the code chunk that actually produces the histogram
# that appears in the slides.  The extra graphical parameters are 
# necessary due to the adjusted resolution of the new displays
par(mai = c(1.5,2,2,2))

hist(survey$TVhours, col="lightgray", 
       cex.lab = 3, cex.axis = 2.5, cex.main = 3.5)
```
</center>

###Indexing multiple columns

```{r}
head(survey[, c(1,5)]) # Data from 1st and 5th columns
head(survey[c("Program", "Editor")]) # Data from "Program" and "Editor"
```

###Indexing rows and columns
- Data frames have two dimensions to index across 

```{r}
survey[6,] # 6th row
survey[6,5] # row 6, column 5
survey[6, "Program"] # Program of 6th survey respondent 
survey[["Program"]][6]  # Program of 6th survey respondent 
```

###More indexing
- In Lab 1, you were introduced to the colon operator `:`

- We can use this operator for indexing

```{r}
survey[1:3,]  # equivalent to head(survey, 3)
survey[3:5, c(1,5)] 
```

###Subsets of data

- We are often interested in learning something a specific subset of the data

```{r, eval = FALSE}
survey[survey$Program=="MISM", ] # Data from the MISM students
survey[which(survey$Program=="MISM"), ] # Does the same thing
```

```{r, echo = FALSE}
survey[survey$Program=="MISM", ] # Data from the MISM students
```

###More subset examples

- Let's pull all of the PPM students who have never used R before

```{r}
survey[survey$Program=="PPM" & survey$Rexperience=="Never used", ]
```

###Cleaner subsetting

- When the subset conditions get long or messy, it is preferable to use the **`subset()`** function

- Here's an example of selecting the OperatingSystem and TVhours responses from all of the students who are either in PPM or Other and who listed their R experience as "Basic competence".

```{r}
subset(survey, select=c("OperatingSystem", "TVhours"),  subset=(Program == "PPM" | Program == "Other") & Rexperience == "Basic competence")
```

###Splitting a long function call

- As your function calls get longer and more complicated, you may find it useful to split them over multiple lines

- Here's one way to rewrite the previous line

```{r}
subset(survey, 
       select = c("OperatingSystem", "TVhours"),  
       subset = (Program == "PPM" | Program == "Other") & 
                (Rexperience == "Basic competence")
       )
```

###Some simple calculations

```{r}
mean(survey$TVhours[survey$Program == "PPM"]) # Average time PPM's spent watching TV
mean(survey$TVhours[survey$Program == "MISM"]) # Average time MISM's spent watching TV 
mean(survey$TVhours[survey$Program == "Other"]) # Average time "Others" spent watching TV
```

- Later on we'll learn a better way of doing these types of calculations by using the **aggregate()** function.

###Defining variables

- If we wanted to focus on a particular column of the data frame, we could always define it as a new variable

```{r}
tv.hours <- survey$TVhours  # Vector of TVhours watched
mean(tv.hours)              # Average time spent watching TV
sd(tv.hours)                # Standard deviation of TV watching time
```

```{r, eval = FALSE}
tv.hours >= 5         # (Settings adjusted to print first 40 elements)
```
```{r, echo = FALSE}
head(tv.hours >= 5, 40)
```

```{r}
sum(tv.hours >= 5)   # How many people watched 5 or more hours of TV?
```

###R coding style

- Coding style (and code commenting) will become increasingly more important as we get into more advanced and involved programming tasks

- A few R "style guides" exist:
    - [Google's](https://google.github.io/styleguide/Rguide.xml)
    - [Hadley Wickham's](http://r-pkgs.had.co.nz/style.html)

- Borrowing Hadley Wickham's words:
> You don’t have to use my style, but you really should use a **consistent** style.

###R style recommendations

- [Hadley Wickham's](http://r-pkgs.had.co.nz/style.html) guide is short and easy to follow


- We'll revisit the question of coding style several times over the course of the class

###Enforced style: Assignment operator

**Assignment operator**.  USE **`<-`**

```{r, eval=FALSE}
student.names <- c("Eric", "Hao", "Jennifer")  # Good
student.names = c("Eric", "Hao", "Jennifer") # Bad
```

- Note: When specifying function arguments, only `=` is valid

```{r, eval=FALSE}
sort(tv.hours, decreasing=TRUE) # Good
sort(tv.hours, decreasing<-TRUE) # Bad!!
```

###Enforced style: Spacing

- Binary operators should have spaces around them

- Commas should have a space after, but not before (just like in writing)

```{r, eval=FALSE}
3 * 4 # Good
3*4 # Bad
which(student.names == "Eric") # Good
which(student.names=="Eric") # Bad
```

- For specifying arguments, spacing around `=` is optional

```{r, eval=FALSE}
sort(tv.hours, decreasing=TRUE) # Accepted
sort(tv.hours, decreasing = FALSE) # Accepted
```

###Enforced style: Variable names

- To make code easy to read, debug, and maintain, you should use **concise** but **descriptive** variable names

- Terms in variable names should be separated by `_` or `.`

```{r, eval=FALSE}
# Accepted
day_one   day.one   day_1   day.1   day1

# Bad
d1   DayOne   dayone   

# Can be made more concise:
first.day.of.the.month
```

- Avoid using variable names that are already pre-defined in R
```{r, eval=FALSE}
# EXTREMELY bad:
c   T   pi   sum   mean   
```
 
###Assignments

- **Homework 1** is posted
  - **Due: Thursday, January 26, 1:20pm**
  - Submit your .Rmd file on Blackboard
  - Course website contains grading rubric

- **Lab 2** is now available
  - Submit Lab 2 on Blackboard by **11:59pm tonight**
  - Remember: To earn a participation point for today's class, you must submit the lab